Measuring matter effects in Neutron star - Black hole binaries with Advanced LIGO

### 2016-04-01 Paper/paper.tex
cleaning up text in S5

% 
\begin{figure}
\centering    
\includegraphics[width=1.\columnwidth]{plots/LambdalErrorBars_vs_Lambda_N49_Log.pdf}
\caption{This figure shows the relative uncertainty in the measurement of $\lambdans$,
as a function of $\lambdans$ itself. The drawn curves are cumulative, in the sense 
that the $N=n$ curve includes information from all of the first $n$ observed events.
Those corresponding to the first $N\leq 5$ events are shown in fading shades of red,
with the ones for higher value of $N$ being shown in grey (going from light to dark in
proportion to $N$).
Green dash-dotted lines are contours of $y=x^{-(1+1/5)}$.
% 
\prayush{We note that the measurement uncertainty from a population follows the
equation for the green lines.}
% 
}
\label{fig:TT_Lambda_vs_Lambda_L500_2000_CI90_0_AllInOne}
\end{figure}
%

% 
\subsection{Multiple identical sources at low SNR}\label{s2:identical_multiple}
% 
% 
Here we show a crude calculation which tells us that approximately $8$ (or $16$) observations
of NSBH signals with $\rho\geq 10$ distributed uniformly in effective volume, would
give us similarly accurate a measurement of $\lambdans$ as would a single detection
with $\rho=50$ (or $\rho=70$).

We want to distribute signals in effective distance $\deff$, which is a combination of 
luminosity distance, inclination angle of the binary's orbital angular momentum
with respect to the detector, and its various sky angles. The signal strength 
scales inversely with $\deff$, if it comprises only of the dominant $(2,2)$ mode, allowing
for various angle-dependent factors to scale out beside the luminosity distance.
% 
In order to consider multiple events, lets imagine a population uniformly distributed
in effective volume, i.e. within a sphere of radius $\deff^\mathrm{max}$. The 
radius of this sphere is set by the lowest SNR that is distinguishable from noise
by LIGO searches. Now, dividie the sphere into $I$ shells of equal width. The radius
of the $i$-th sphere would then be $D_i = \deff^\mathrm{max} (i - 1/2)/I$. The 
measurement uncertainty in $\lambdans$ scales inversely with the SNR, and hence
directly with the effective distance to the source. Therefore, if we have a 
measurement error $\sigma_0$ for a source located at $\deff = D_0$, the same error
for the same source located within the $i$-th shell would be 
$$
\sigma_i = \sigma_0 \frac{D_i}{D_0}.
$$
Independent measurements of $\lambdans$ for identical sources at different distances
would have their combined error given by
\begin{equation}\label{eq:1oversigma}
\frac{1}{\sigma^2} = \sum_{i=1}^I \frac{N_i}{\sigma_i^2} = \left(\frac{D_0}{\sigma_0}\right)^2 \sum_{i=1}^I\frac{N_i}{D_i^2},
\end{equation}
where $N_i$ is the number of sources detected in the $i$-th shell, and is a random
variable, with its probability proportional to the volume of the shell, i.e.
$$
p(N_i) \propto \frac{V_i}{V_\mathrm{total}} \propto \dfrac{\left(\deff^\mathrm{max} \frac{i}{I}\right)^3 - \left(\deff^\mathrm{max} \frac{i-1}{I}\right)^3}{(\deff^\mathrm{max})^3} \propto \frac{1}{I^3} [i^3 - (i-1)^3].
$$
Therefore, the expected number of detections in the $i$-th shell would be
$$
\langle N_i\rangle = \frac{N}{I^3} [i^3 - (i-1)^3],
$$
where $N=\sum_{i=1}^I N_i$ is the total number of NSBH detections. $N$ is expected
to be Poisson distributed around the mean detection rate
$\mathcal{R}\equiv\langle N\rangle$, where $\mathcal{R}$ can vary from $0.6-1000$ per
$\mathrm{Gpc}^3$ per year~\cite{Abadie:2010cfa}. If there are $n$ resulting detections
a year per unit effective volume, then
\begin{equation}
\langle N\rangle = \int_0^{\deff^\mathrm{max}} 4\pi n D^2 \D D = \frac{4\pi}{3} n (\deff^\mathrm{max})^3,
\end{equation}
and
\begin{eqnarray}
 \langle \frac{1}{\sigma^2}\rangle &=& \left(\frac{D_0}{\sigma_0}\right)^2 \int_0^{\deff^\mathrm{max}} \frac{4\pi n D^2 }{D^2}\D D\\
 &=& \left(\frac{D_0}{\sigma_0}\right)^2 4\pi n \deff^\mathrm{max},
\end{eqnarray}
where in the previous equation we have converted the summation in Eq.~\ref{eq:1oversigma} 
to an integral. So the root-mean-square averaged measurement error from $N$ sources 
distributed uniformly in effective volume within a sphere of radius $\deff^\mathrm{max}$
would be
\begin{equation}\label{eq:rmsSigmaIdenticalSources}
 \sigma_{avg} \equiv \frac{1}{\sqrt{\langle1/\sigma^{2}\rangle}} = \frac{\sigma_0}{D_0} \deff^\mathrm{max} \frac{1}{\sqrt{3\langle N\rangle}}.
\end{equation}
Let us analyze the implications of Eq.~\ref{eq:rmsSigmaIdenticalSources}. It is
straightforward to deduce that the same measurement certainty as afforded by a single
observation with a high SNR $\alpha$ can be obtained from $N = \alpha^2/300$ observations
uniformly distributed in effective volume with SNRs $\geq 10$. E.g., to get to the level 
of certainty afforded at SNR$=70$, we would need $49/3\approx 16-17$ realistic detections.

There are two main caveats to this rudimentary calculation, which was outlined in 
Ref.~\cite{Markakis:2010mp}, (i) it applies to identical sources,
which is astrophysically next to impossible to achieve exactly, but could perhaps happen 
approximately; (ii) the average measurement error $\sigma_{avg}$ has been defined as in 
the first equality in Eq.~\ref{eq:rmsSigmaIdenticalSources}.

% 
\subsection{Astrophysical source population}\label{s2:astro_multiple}
% 
Lets try a respectable Bayesian approach. Imagine we have $N$ stretches of data,
$d_1, d_2, \cdots, d_N$, each containing a single observation of an NSBH binary. Let $K$
denote our collective prior knowledge, except for expectations on binary parameters
themselves, which will enter the following calculation explicitly. Let the gravitational wave
signal from each binary be characterized by the parameters 
$\vec{\theta} = \{\mbh, \mns, \chibh, \chins, \rho\}$ and $\lambdans$. Note that we have folded
in the luminosity distance, inclination angle of the binary's orbital angular momentum with
respect to the plane of the detector, and its sky location angles into a single parameter $\rho$,
which is the SNR of the received signal. Note also that $\vec{\theta}$ does not include
$\lambdans$ as the latter would be treated independently. In addition, we assume
that all observed NS's have the same equation of state, that uniquely determines their
$\lambdans$ parameter. In other words, we assume that we know its mass perfectly.
% 
Using Bayes' theorem, the measured probability distribution function for the $\lambdans$
parameter is given as
\begin{equation}\label{eq:lambdaMultiple}
 p(\lambdans | d_1, d_2, \cdots, d_N; K) = p(\lambdans | K)^{\alpha}\prod_{i=1}^N p(\lambdans | d_i, K),
\end{equation}
where $p(\lambdans|K)$ is the prior expectation for $\lambdans$, and we have assumed
that all of the $N$ observations are independent. [Note: $\alpha$ will be calculated 
soon, it should be a function only of N.]. A priori, we will assume that no particular value
of $\lambdans$ is more preferred over another, and $\lambdans\leq 4000$, ie 
$$
p(\lambdans | K) = \mathrm{Rect}\left(\frac{\lambdans-2000}{4000}\right)
$$
% 
The second set of terms $p(\lambdans | d_i, K)$ are the inferred posterior distribution
functions for $\lambdans$ from \textit{each} observation independently, and are 
calculated by marginalizing over all other binary parameters, i.e.
\begin{equation}
 p(\lambdans | d_n, K) = \int \D\, \vec{\theta} p(\vec{\theta}, \lambdans | d_n, K).
\end{equation}
The joint inferred probability distribution of all parameters $\vec{\theta}\cup\{\lambdans\}$ 
is given by
\begin{equation}
 p(\vec{\theta}, \lambdans | d_n, K) = \dfrac{p(d_n|\vec{\theta}, \lambdans, K)\,p(\vec{\theta}, \lambdans | K)}{p(d_n|K)}.
\end{equation}
Here, $p(\vec{\theta}, \lambdans | K)$ is the prior probability for binary parameters of taking 
particular values. In this paper, we use a uniform prior on the individual masses within 
pre-chosen ranges, as well as a uniform prior on the spin of the black hole. We set neutron star's
spin identically $=0$, and its tidal deformability parameter $\lambdans$ to be equally likely
between $[0, 4000]$. The prior probability of obtaining a particular realization of data
$p(d_n|K)$ is absorbed into the overall normalization. Finally, the first term in the numerator
$p(d_n|\vec{\theta}, \lambdans, K)$ is the likelihood of obtaining the given stretch of 
data $d_n$ if we assume that a signal parameterized by $\vec{\theta}\cup\{\lambdans\}$ is buried
in it, i.e.
\begin{equation}\label{eq:likelihood}
 \LL(\vec{\theta}, \lambdans) \equiv p(\vec{\theta}, \lambdans | K) = \mathcal{N} \mathrm{exp}[- \langle d_n - h | d_n - h\rangle ],
\end{equation}
where $h\equiv h(\vec{\theta}, \lambdans)$ is our waveform model, and the detector noise
weighted inner-product $\langle a|b\rangle$ is defined as 
$$
\langle a|b\rangle \equiv 4\,\mathrm{Re}\left[\int_0^\infty \dfrac{\tilde{a}(f) \tilde(b)(f)^*}{S_n(|f|)}\,\D f\right],
$$
where $\tilde{a}$ means the Fourier transform of the finite time series $a(t)$, and 
$S_n(|f|)$ is the one-sided amplitude spectrum of detector noise. The definition of $\LL$ in 
Eq.~\ref{eq:likelihood} assumes that the instrument noise is colored Gaussian.

>>>>

% 
\begin{figure}
\centering    
\includegraphics[width=1.\columnwidth]{plots/LambdalErrorBars_vs_Lambda_N49_Log.pdf}
\caption{This figure shows the relative uncertainty in the measurement of $\lambdans$,
as a function of $\lambdans$ itself. The drawn curves are cumulative, in the sense 
that the $N=n$ curve includes information from all of the first $n$ observed events.
Those corresponding to the first $N\leq 5$ events are shown in fading shades of red,
with the ones for higher value of $N$ being shown in grey (going from light to dark in
proportion to $N$).
Green dash-dotted lines are contours of $y=x^{-(1+1/5)}$.
% 
\prayush{We note that the measurement uncertainty from a population follows the
equation for the green lines.}
% 
}
\label{fig:TT_Lambda_vs_Lambda_L500_2000_CI90_0_AllInOne}
\end{figure}
%
% 
\textbf{Multiple identical sources at low SNR: }\label{s2:identical_multiple}
% 
% 
First we show a crude calculation which tells us that approximately $8$ (or $16$) 
observations
of NSBH signals with $\rho\geq 10$ distributed uniformly in effective volume, would
give us similarly accurate a measurement of $\lambdans$ as would a single detection
with $\rho=50$ (or $\rho=70$).

We want to distribute signals in effective distance $\deff$, which is a combination of 
luminosity distance, inclination angle of the binary's orbital angular momentum
with respect to the detector, and its various sky angles. The signal strength 
scales inversely with $\deff$, if it comprises only of the dominant $(2,2)$ mode, allowing
for various angle-dependent factors to scale out beside the luminosity distance.
% 
In order to consider multiple events, lets imagine a population uniformly distributed
in effective volume, i.e. within a sphere of radius $\deff^\mathrm{max}$. The 
radius of this sphere is set by the lowest SNR that is distinguishable from noise
by LIGO searches. Now, divide the sphere into $I$ shells of equal width. The radius
of the $i$-th sphere would then be $D_i = \deff^\mathrm{max} (i - 1/2)/I$. The 
measurement uncertainty in $\lambdans$ scales inversely with the SNR, and hence
directly with the effective distance to the source. Therefore, if we have a 
measurement error $\sigma_0$ for a source located at $\deff = D_0$, the same error
for the same source located within the $i$-th shell would be 
$\sigma_i = \sigma_0 \dfrac{D_i}{D_0}.$
Independent measurements of $\lambdans$ for identical sources at different distances
would have their combined error given by
\begin{equation}\label{eq:1oversigma}
\frac{1}{\sigma^2} = \sum_{i=1}^I \frac{N_i}{\sigma_i^2} = \left(\frac{D_0}{\sigma_0}\right)^2 \sum_{i=1}^I\frac{N_i}{D_i^2},
\end{equation}
where $N_i$ is the number of sources detected in the $i$-th shell.
% , and is a random
% variable, with its probability proportional to the volume of the shell, i.e.
% $$
% p(N_i) \propto \frac{V_i}{V_\mathrm{total}} \propto \dfrac{\left(\deff^\mathrm{max} \frac{i}{I}\right)^3 - \left(\deff^\mathrm{max} \frac{i-1}{I}\right)^3}{(\deff^\mathrm{max})^3} \propto \frac{1}{I^3} [i^3 - (i-1)^3].
% $$
% Therefore, the expected number of detections in the $i$-th shell would be
% $$
% \langle N_i\rangle = \frac{N}{I^3} [i^3 - (i-1)^3],
% $$
% where $N=\sum_{i=1}^I N_i$ is the total number of NSBH detections. $N$ is expected
% to be Poisson distributed around the mean detection rate
% $\mathcal{R}\equiv\langle N\rangle$, where $\mathcal{R}$ can vary from $0.6-1000$ per
% $\mathrm{Gpc}^3$ per year~\cite{Abadie:2010cfa}. If there are $n$ resulting detections
% a year per unit effective volume, then
% \begin{equation}
% \langle N\rangle = \int_0^{\deff^\mathrm{max}} 4\pi n D^2 \D D = \frac{4\pi}{3} n (\deff^\mathrm{max})^3,
% \end{equation}
% and
% \begin{eqnarray}
%  \langle \frac{1}{\sigma^2}\rangle &=& \left(\frac{D_0}{\sigma_0}\right)^2 \int_0^{\deff^\mathrm{max}} \frac{4\pi n D^2 }{D^2}\D D\\
%  &=& \left(\frac{D_0}{\sigma_0}\right)^2 4\pi n \deff^\mathrm{max},
% \end{eqnarray}
% where in the previous equation we have converted the summation in Eq.~\ref{eq:1oversigma} 
% to an integral. 
Continuing on this line of inquiry, Ref.~\cite{Markakis:2010mp} calculates the 
root-mean-square averaged measurement error from 
$\langle N\rangle:=\langle\sum N_i\rangle$ sources distributed uniformly in
effective volume within a sphere of radius $\deff^\mathrm{max}$ to be
\begin{equation}\label{eq:rmsSigmaIdenticalSources}
 \sigma_{avg} := \frac{1}{\sqrt{\langle1/\sigma^{2}\rangle}} = \frac{\sigma_0}{D_0} \deff^\mathrm{max} \frac{1}{\sqrt{3\langle N\rangle}},
\end{equation}
given a fiducial pair $(\sigma_0, D_0)$.
Further, it is straightforward to deduce from Eq.~\ref{eq:rmsSigmaIdenticalSources}
that the same measurement certainty as afforded by a single
observation with a high SNR $\alpha$ can be obtained from $N = \alpha^2/300$ observations
uniformly distributed in effective volume with SNRs $\geq 10$. E.g., to get to the level 
of certainty afforded at SNR$=70$, we would need $49/3\approx 16-17$ realistic detections.

There are two main caveats to this rudimentary calculation, which was outlined in 
Ref.~\cite{Markakis:2010mp}, (i) it applies to identical sources,
which is astrophysically next to impossible to achieve exactly, but could perhaps happen 
approximately; (ii) the combined average measurement error $\sigma_{avg}$ has been defined
as in the first equality in Eq.~\ref{eq:rmsSigmaIdenticalSources}. To overcome both,
we next perform a Bayesian population analysis.

\textbf{Astrophysical source population: }\label{s2:astro_multiple}
% 
Imagine we have $N$ stretches of data,
$d_1, d_2, \cdots, d_N$, each containing a single observation of an NSBH binary. Let $K$
denote our collective prior knowledge, except for expectations on binary parameters
themselves, which will enter the following calculation explicitly. Let the gravitational wave
signal from each binary be characterized by the parameters 
$\vec{\theta} = \{\mbh, \mns, \chibh, \chins, \rho\}$ and $\lambdans$. Note that we have folded
in the luminosity distance, inclination angle of the binary's orbital angular momentum with
respect to the plane of the detector, and its sky location angles into a single parameter $\rho$,
which is the SNR of the received signal. Note also that $\vec{\theta}$ does not include
$\lambdans$ as the latter would be treated independently. In addition, we assume
that all observed NS's have the same equation of state, that uniquely determines their
$\lambdans$ parameter. In other words, we assume that we know its mass perfectly.
% 
Using Bayes' theorem, the measured probability distribution function for the $\lambdans$
parameter is given as
\begin{equation}\label{eq:lambdaMultiple}
 p(\lambdans | d_1, d_2, \cdots, d_N; K) = p(\lambdans | K)^{\alpha}\prod_{i=1}^N p(\lambdans | d_i, K),
\end{equation}
where $p(\lambdans|K)$ is the prior expectation for $\lambdans$, and we have assumed
that all of the $N$ observations are independent. [Note: $\alpha$ will be calculated 
soon, it should be a function only of N.]. A priori, we will assume that no particular value
of $\lambdans$ is more preferred over another, and $\lambdans\leq 4000$, ie 
$$
p(\lambdans | K) = \mathrm{Rect}\left(\frac{\lambdans-2000}{4000}\right)
$$
% 
The second set of terms $p(\lambdans | d_i, K)$ are the inferred posterior distribution
functions for $\lambdans$ from \textit{each} observation independently, and are 
calculated by marginalizing over all other binary parameters, i.e.
\begin{equation}
 p(\lambdans | d_n, K) = \int \D\, \vec{\theta} p(\vec{\theta}, \lambdans | d_n, K).
\end{equation}
The joint inferred probability distribution of all parameters $\vec{\theta}\cup\{\lambdans\}$ 
is given by
\begin{equation}
 p(\vec{\theta}, \lambdans | d_n, K) = \dfrac{p(d_n|\vec{\theta}, \lambdans, K)\,p(\vec{\theta}, \lambdans | K)}{p(d_n|K)}.
\end{equation}
Here, $p(\vec{\theta}, \lambdans | K)$ is the prior probability for binary parameters of taking 
particular values. In this paper, we use a uniform prior on the individual masses within 
pre-chosen ranges, as well as a uniform prior on the spin of the black hole. We set neutron star's
spin identically $=0$, and its tidal deformability parameter $\lambdans$ to be equally likely
between $[0, 4000]$. The prior probability of obtaining a particular realization of data
$p(d_n|K)$ is absorbed into the overall normalization. Finally, the first term in the numerator
$p(d_n|\vec{\theta}, \lambdans, K)$ is the likelihood of obtaining the given stretch of 
data $d_n$ if we assume that a signal parameterized by $\vec{\theta}\cup\{\lambdans\}$ is buried
in it, i.e.
\begin{equation}\label{eq:likelihood}
 \LL(\vec{\theta}, \lambdans) \equiv p(\vec{\theta}, \lambdans | K) = \mathcal{N} \mathrm{exp}[- \langle d_n - h | d_n - h\rangle ],
\end{equation}
where $h\equiv h(\vec{\theta}, \lambdans)$ is our waveform model, and the detector noise
weighted inner-product $\langle a|b\rangle$ is defined as 
$$
\langle a|b\rangle \equiv 4\,\mathrm{Re}\left[\int_0^\infty \dfrac{\tilde{a}(f) \tilde(b)(f)^*}{S_n(|f|)}\,\D f\right],
$$
where $\tilde{a}$ means the Fourier transform of the finite time series $a(t)$, and 
$S_n(|f|)$ is the one-sided amplitude spectrum of detector noise. The definition of $\LL$ in 
Eq.~\ref{eq:likelihood} assumes that the instrument noise is colored Gaussian.
